---
title: "STAT_4870_Project"
author: "Waleed Abdulla"
date: "2025-04-17"
output: html_document
---

#Simple/Multiple Lin Reg doesn't count but can be used for comparison
#PCA
#AIC
#BIC
#Log Regression
#QDA/LDA
#kNN

New Libraries: fastDummies

Libraries
```{r, include=FALSE}
library(fpp3)
library(fastDummies)
```


```{r}
Bank <- read.csv("C:/Users/13147/Downloads/bank-full.csv", sep = ";")
Bank %>% head(3)
```

Step 1: Literature review and theoretical model - What question you are trying to answer?

 - Goal: Predict if the client will subscribe to a term deposit (variable y).
 - Goal: Find the key predictors 
 - Goal: Create a good applied regression model.
 
 
Step 2: Clean our data.

  We need to create indicator variables for our categorical predictors
  
```{r}
Bank %>% select_if(is.character) %>% head(3)
```
There are 10 categorical indicators. Luckily, the fastDummies package automatically creates dummy variables for each categorical indicators.

```{r}
BankN <- Bank %>% dummy_cols() %>% select(-c("job","marital","education","default","housing","loan","contact",
                                    "month","poutcome","y_no","y_yes")) %>% as.data.frame()
str(BankN)
```

Now, if we want to use a categorical variable, we'll use a preset filter that locates the relevant columns. We can even create a matrix. For example, say we want to use the *job* column. 

```{r}
BankN %>% select(starts_with("job")) %>% head()
BankN %>% select(starts_with("job")) %>% as.matrix() %>% head()
```
We need to fix the binary categorical variables to be in one column instead of two.

```{r}
BankN <- BankN %>% rename(default = default_yes, housingLoan = housing_yes, personalLoan = loan_yes) %>%
  select(-c(default_no, housing_no, loan_no))
```

The *campaign* column is the number of contacts performed *during* this campaign and the *previous* column is the number of contacts performed *before* this campaign. Let's combine both to get the total number of contacts

```{r}
BankN <- BankN %>% mutate(contacts = campaign + previous) %>% select(-c("campaign","previous"))
```

The *day* column and the *month* column are linked, so we can remove the *day* column all together.

```{r}
BankN <- BankN %>% select(-day)
```

Final dataset (for now)
```{r}
BankN %>% head()
```




---- IGNORE ----

Before checking for assumptions, let's create a quick, simple linear regression and check the plot

```{r}

```

 
Step 2: Make sure our data meets the assumptions.
  For now, lets check the four main assumptions for *linear regression*
  
  First Assumption: Independence of Observations

```{r}

```

  Second Assumption: Normality
  
```{r}

```
---- IGNORE ----